---
title: "Genetic Correlation"
author: "<h4>Author: <i>Brian M. Schilder</i></h4>" 
date: "<h4>Most recent update: <i>`r format( Sys.Date(), '%b-%d-%Y')`</i></h4>"
output:
  BiocStyle::html_document:
vignette: >
    %\VignetteIndexEntry{phenomix} 
    %\usepackage[utf8]{inputenc}
    %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=TRUE, message=FALSE} 
library(phenomix) 
```


# Import data

## obj 

```{r}
obj <- phenomix::get_HPO()
obj <- obj[seq(100),]
```


# Compute correlations

## Full matrix 

Next, we will compute all pairwise trait-trait correlations.
`extract_cor` returns the object with the correlation matrix 
(within `obj@graphs$cor`).

```{r}
obj <- phenomix::get_cor(obj = obj,
                         return_obj = TRUE)
```

The correlation matrix by itself 
can easily be extracted by setting `return_obj = FALSE`.

```{r}
cmat <- phenomix::get_cor(obj = obj,
                          return_obj = FALSE)
```

### Find neighbors

`find_neighbors` makes finding traits that correlate with your 
traits of interest even easier. You can enter a list of terms to `var1_search`
to perform a case-insensitive substring search to find traits of interest. 

`find_neighbors` will then melt and sort the correlation matrix into a 
data.table for further processing.

```{r}
knn <- phenomix::find_neighbors(obj = obj, 
                                var1_search = c("parkinson","alzheimer"), 
                                label_col = "HPO_label")
```

### Plot

Finally, we can plot the top N most similar traits to your traits of interest.

```{r}
gg_cor <- phenomix::plot_trait_cor(knn = knn,
                                   top_n = 30)
```


## Reductions

By default, correlations are performed on the gene x trait matrix.
However, it is possible to also compute correlations from the 
reduced embeddings matrix stored within the same object.

This approach will be faster than running correlations on the full matrix,
but also tends to be less accurate since some information is thrown out during 
dimensionality reduction.

It's also worth mentioned that if the graph of interest doesn't exist, `find_neighbors` will automatically compute it based on the input arguments.

## Find neighbors

```{r}
knn <- phenomix::find_neighbors(obj = obj, 
                                keys = "pca",
                                var1_search = c("parkinson","alzheimer"), 
                                label_col = "HPO_label")
```

## Plot

```{r}
gg_cor <- phenomix::plot_trait_cor(knn = knn,
                                   top_n = 5)
```



# dPRS

## Import dPRS data

Now let's try with another dataset, this time based on GWAS sumatats: dPRS. 

```{r, eval=FALSE}
dprs <- phenomix::get_dPRS()
dprs <- dprs[seq(100)]
```

## Find neighbors

```{r, eval=FALSE}
knn <- phenomix::find_neighbors(obj = dprs, 
                                var1_search = c("parkinson"), 
                                label_col = "GBE_NAME")
```

## Plot

```{r, eval=FALSE}
gg_cor <- phenomix::plot_trait_cor(knn = knn, 
                                   top_n = 20)
```



# Session Info 

<details> 

```{r Session Info}
utils::sessionInfo()
```

</details>  

